---
title: "w271_ChristianMillsop_TennisonYu_Lab1"
author: "Christian Millsop, Tennison Yu"
date: "January 22, 2019"
header-includes:
  - \usepackage{textcomp}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 4

The failure of an O-ring on the space shuttle Challenger's booster rockets led to its destruction in 1986. Using data on previous space shuttle launches, Dalal et al. (1989) examine the probability of an O-ring failure as a function of temperature at launch and combustion pressure. Data from their paper is included in the challenger.csv file. Below are the variables:
+ Flight: Flight number
+ Temp: Temperature (F) at launch
+ Pressure: Combustion pressure (psi)
+ O.ring: Number of primary field O-ring failures
+ Number: Total number of primary field O-rings (six total, three each for the two booster rockets)

The response variable is O.ring, and the explanatory variables are Temp and Pressure. Complete the following:
+ (a) The authors use logistic regression to estimate the probability an O-ring will fail. In order to use this model, the authors needed to assume that each O-ring is independent for each launch. Discuss why this assumption is necessary and the potential problems with it. Note that a subsequent analysis helped to alleviate the authors' concerns about independence.
  + Rocket motors are potentially re-used.  Are O-rings replaced each time?
  + There are 3 primary O-rings per rocket.  Operational anomalies that occur in a rocket likely affect all o-rings on the rocket.
  + https://stats.stackexchange.com/questions/259704/is-there-i-i-d-assumption-on-logistic-regression

+ (b) Estimate the logistic regression model using the explanatory variables in a linear form.
```{r}
data = read.csv(file="./challenger.csv", header=TRUE)
data$Prob = data$O.ring/data$Number
```

```{r}
fit.a = glm(formula = Prob ~ Temp + Pressure, family=binomial(link=logit),data = data)
summary(fit.a)
```

+ (c) Perform LRTs to judge the importance of the explanatory variables in the model.
```{r}
fit.a.null = glm(formula = Prob ~ 1, family=binomial(link=logit),data = data)
fit.a.reduced = glm(formula = Prob ~ Temp, family=binomial(link=logit),data = data)
anova(fit.a.null, fit.a.reduced, fit.a, test="Chisq")
```

+ (d) The authors chose to remove Pressure from the model based on the LRTs. Based on your results, discuss why you think this was done.  Are there any potential problems with removing this variable?
  + Pressure is less important than Temperature.  Neither explanatory variable is statistically significant by LRT though.
  + Parsimony is desirable.
  + PV = nRT, we expect some relationship or interaction between pressure and temperature.
  + High pressure can cause blow holes, which expose the O-ring to high temperature

## Question 5

Continuing Exercise 4, consider the simplified model $logit(\pi) = \beta_0 + \beta_1 Temp$ where $\pi$ is the probability of an O-ring failure.  Complete the following:
+ (a) Estimate the model
```{r}
fit.b = glm(formula = Prob ~ Temp, family=binomial(link=logit),data = data)
summary(fit.b)
```

+ (b) Construct two plots: (1) $\pi$ vs. $Temp$ and (2) Expected number of failures vs. $Temp$.  Use a temperature range of $31\textdegree$ to $81\textdegree$ on the x-axis even though the minimum temperature in the data set was $53\textdegree$.
  + The first plot, $\pi$ vs. $Temp$, is the probability that any given o-ring will fail at a certain temperature.
  + The second plot is the sum of the probabilities for all 6 of the o-rings to fail at a certain temperature.
```{r}
plotrange = seq(31, 81, 1)
pred_odds = predict(fit.b, newdata=data.frame(Temp=plotrange), type="link")
plot(plotrange, exp(pred_odds)/(1+exp(pred_odds)), xlim=c(31,81), ylab="Probability of failure", xlab="Temperature (F)")
plot(plotrange, exp(pred_odds)/(1+exp(pred_odds))*6, xlim=c(31,81), ylab="Expected number of failures", xlab="Temperature (F)")

```

+ (c) Include the 95% Wald confidence interval bands for $\pi$ on the plot.  Why are the bands much wider for lower temperatures than for higher temperatures?
  + The CI bands are wider at lower temperatures because there is no data below $53\textdegree$.
```{r}
ci.plot = function(model,xdata, alpha) {
  z = qnorm(1-alpha/2)
  predictions = predict(model, newdata=data.frame(Temp=xdata), type="link", se.fit=TRUE)
  lower_ci = predictions$fit - z*predictions$se.fit
  upper_ci = predictions$fit + z*predictions$se.fit
  list(lower=exp(lower_ci)/(1+exp(lower_ci)),upper=exp(upper_ci)/(1+exp(upper_ci)))
}

plot(data$Temp, data$Prob, xlim = c (31, 81), ylim=c(0,1), xlab="Temperature (F)", ylab="Probability of failure")
curve (expr = predict (object = fit.b, newdata = data.frame(Temp = x), type = "response"), col = "blue", add = TRUE, xlim = c (31, 81), ylim=c(0,1))
curve (expr = ci.plot(fit.b,x,0.05)$lower, col = "red", add = TRUE, xlim = c (31, 81), ylim=c(0,1))
curve (expr = ci.plot(fit.b,x,0.05)$upper, col = "red", add = TRUE, xlim = c (31, 81), ylim=c(0,1))
```

+ (d) The temperature was $31\textdegree$ at launch for the Challenger in 1986.  Estimate the probability of an O-ring failure using this temperature, and compute a corresponding confidence interval.  Discuss what assumptions need to be made in order to apply the inference procedures.
```{r}
predicted_odds = predict(fit.b, newdata=data.frame(Temp=31), type="link", se.fit = TRUE)
predicted_prob = exp(predicted_odds$fit)/(1+exp(predicted_odds$fit))
ci_odds = c(predicted_odds$fit - qnorm(1-.05/2)*predicted_odds$se.fit, predicted_odds$fit + qnorm(1-.05/2)*predicted_odds$se.fit)
ci = exp(ci_odds)/(1+exp(ci_odds))
"Predicted probability of at least one O-ring failure with 95% confidence interval:"
paste(round(ci[1],5), "<", round(predicted_prob,5), "<", round(ci[2],5))
```

+ (e) Rather than using Wald or profile LR intervals for the probability of failure, Dalal et al. (1989) use a parameteric bootstrap to compute intervals.  Their process was to (1) simulate a large number of data sets ($n = 23$ for each) from the estimated model of $logit(\pi) = \beta_0 + \beta_1 Temp$; (2) estimate new models for each data set, say $logit(\hat \pi^*) = \hat \beta_0^* + \hat \beta_1^* Temp$; and (3) compute $\hat \pi^*$ at a specific temperature of interest.  The authors used the 0.05 and 0.95 observed quantiles from the $\hat \pi^*$ simulated distribution as their 90% confidence interval limits.  Using the parametric bootstrap, compute 90% confidence intervals separately at temperatures of $31\textdegree$ and $72\textdegree$.
```{r}
bootstrap =function(test_values){
  samples = data[sample(nrow(data), size=23, replace=TRUE), ]
  fit.boot = suppressWarnings(glm(Prob ~ Temp, family=binomial(link=logit),data = samples, control = list(maxit = 50)))
  predicted_odds = predict(fit.boot, newdata=data.frame(Temp=test_values), type="link")
  predicted_prob = exp(predicted_odds)/(1+exp(predicted_odds))
}
x = replicate(10000,bootstrap(c(31,72)))
xt = aperm(x,c(2,1))

paste("90% CI for 31F:")
quantile(xt[,1], probs=c(.05,.95))
paste("90% CI for 72F:")
quantile(xt[,2], probs=c(.05, .95))


```


+ (f) Determine if a quadratic term is needed in the model for the temperature.
  + Adding a quadratic term for temperature didn't improve the model very much.
  + The residual deviance decreased slightly from 3.01 to 2.93 (~3%).
  + The AIC improved from 7.2 to 9.2
  + LRT doesn't indicate that temperature^2 is important
  + Is there a physical reason why we would suspect temperature is quadratic?
```{r}
library(car)
fit.c = glm(formula = Prob ~ Temp + I(Temp^2), family=binomial(link=logit),data = data)
summary(fit.c)
Anova(fit.c, test="LR")
anova(fit.b, fit.c, test="Chisq")
```

3. In addition to the questions in Question 4 and 5, answer the following questions:

    a. Interpret the main result of your final model in terms of both odds and probability of failure 
      + Odds: The coefficient for temperature is -0.1156.  The odds of failure decrease by $\exp (c \beta_1)$ for each unit increase in c.  For example, a 1 degree increase in temperature yields a 0.89 decrease in the odds of failure.
      + Probability:  The probability of failure are $\hat \pi = \frac{\exp (-0.1156 Temp + 5.0850)}{1 + \exp (-0.1156 Temp + 5.0850)}$.  Increase in temperature results in reduced probability of failure.
        + Does this need a confidence interval?
      

    b. With the same set of explanatory variables in your final model, estimate a linear regression model. Explain the model results; conduct model diagnostic; and assess the validity of the model assumptions.  Would you use the linear regression model or binary logistic regression in this case.  Please explain.




